---
title: "Multi-task learning from single-cell multimodal omics with matilda"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Getting started with matilda}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

[Matilda](https://github.com/PYangLab/Matilda) is a multi-task framework for
single-cell multimodal omics (RNA, ADT, ATAC). A single multimodal variational
autoencoder plus classifier is trained once and then reused for five tasks:
**classification**, **dimension reduction**, **feature selection**,
**data simulation**, and training-time **augmentation**.

This package runs the original Matilda implementation from R on
`SingleCellExperiment` / `MultiAssayExperiment` objects, writing results back
into the object — so it slots into a standard Bioconductor workflow. The Python
runtime is provisioned automatically by `r BiocStyle::Biocpkg("basilisk")`; you
never install or manage Python yourself. A GPU is used when available and the
package falls back to CPU otherwise.

# Input

A multimodal experiment is a `SingleCellExperiment` with RNA in the main assay
and the other modalities as `altExps`:

```{r data}
library(matilda)
library(SingleCellExperiment)

# A small synthetic multimodal example: RNA counts in the main assay, ADT and
# ATAC as altExps, and a colData column "cell_type" with the labels.
sce <- matilda_example_sce()
sce
```

Matilda expects raw counts for RNA/ADT and the Seurat gene-activity score for
ATAC.

# Train once, run many tasks

The trained model is stored inside the object, so the workflow is pipeable and
each task writes its results back:

```{r workflow, eval = FALSE}
sce <- matilda_train(sce, label = "cell_type")   # model stored in metadata(sce)$matilda
sce <- matilda_reduce(sce)                         # reducedDim(sce, "MATILDA")
sce <- matilda_classify(sce)                       # colData$matilda_pred / $matilda_prob

reducedDim(sce, "MATILDA")[1:5, 1:3]
table(sce$matilda_pred)
```

Feature selection returns per-cell-type marker importance, and simulation
returns a new `SingleCellExperiment` of synthetic cells:

```{r tasks, eval = FALSE}
markers <- matilda_markers(sce, method = "IntegratedGradient")
head(markers)                                      # celltype / feature / importance

sim <- matilda_simulate(sce, celltype = "B.Naive", n = 200)
```

# Reference and query

To apply a trained reference model to new (query) cells, pass it via `reference`:

```{r query, eval = FALSE}
query <- matilda_classify(query, reference = sce)
query <- matilda_reduce(query,  reference = sce)
```

# Without SingleCellExperiment

Plain matrices also work; training then returns a model handle and tasks return
lists:

```{r matrices, eval = FALSE}
fit <- matilda_train(rna = rna_mat, adt = adt_mat, atac = atac_mat, cty = labels)
res <- matilda_classify(list(rna = q_rna, adt = q_adt, atac = q_atac), reference = fit)
res$pred; res$prob
```

# Notes

- The first call builds the bundled Python environment (one-off, then cached).
- Set `device = "cpu"` to force CPU, or `"cuda"` to require a GPU.
- `matilda_example_teaseq()` returns paths to the bundled TEA-seq demo.

# Session info

```{r sessioninfo}
sessionInfo()
```
