---
title: "Matilda tutorial 0 — Preparing your data"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Before training, your data has to be in a shape Matilda accepts. This tutorial
shows the **input contract** and how to reach it from the formats people usually
start with (`.h5ad`, 10x `.h5`, Seurat, plain matrices, Matilda `.h5`).

## What Matilda expects

The primary input is a **`SingleCellExperiment`** (Matilda also accepts
`MultiAssayExperiment`, `Seurat`, or plain matrices):

- **RNA** in the main assay — **raw counts**.
- **ADT** in `altExp(sce, "ADT")` — **raw counts** (optional).
- **ATAC** in `altExp(sce, "ATAC")` — **gene-activity score**, gene-level (optional).
- **cell-type labels** in a `colData` column, e.g. `sce$cell_type`.
- every modality shares the **same cells (columns), in the same order**.
- assays are oriented **features × cells** (genes in rows).

(RNA + at least one of ADT/ATAC = multimodal; RNA-only also works.)

```{r target}
source("_setup.R")
sce <- matilda_example_sce()      # <- this is the shape you are aiming for
sce
SummarizedExperiment::assayNames(sce)
SingleCellExperiment::altExpNames(sce)
head(sce$cell_type)
```

> **Data semantics matter:** RNA/ADT must be **raw counts** (not log/normalised),
> and ATAC must be a **gene-activity score** (e.g. `Signac::GeneActivity()`), *not*
> raw peak counts. Matilda does its own library-size + log2 + z-score internally.

## From `.h5ad` (AnnData / scanpy)

One modality per file → read each with `zellkonverter`, then assemble. (`readH5AD`
returns a genes × cells SCE, transposing AnnData's cells × genes for you.)

```{r h5ad, eval = FALSE}
# BiocManager::install("zellkonverter")
library(zellkonverter); library(SingleCellExperiment)

rna  <- readH5AD("rna.h5ad")
adt  <- readH5AD("adt.h5ad")
atac <- readH5AD("atac_geneactivity.h5ad")

# Make sure the RAW COUNTS are the main assay named "counts"
# (readH5AD puts AnnData's X in assay "X"; raw counts may be in a layer):
assayNames(rna)                                  # inspect first
assay(rna, "counts") <- assay(rna, "X")          # or assay(rna, "<your raw layer>")

# Align cells across modalities, then attach ADT/ATAC as altExps:
common <- Reduce(intersect, list(colnames(rna), colnames(adt), colnames(atac)))
rna  <- rna[, common]; adt <- adt[, common]; atac <- atac[, common]
altExp(rna, "ADT")  <- adt
altExp(rna, "ATAC") <- atac
rna$cell_type <- colData(rna)$cell_type          # your label column
sce <- rna
```

## From 10x CellRanger `.h5` (CITE-seq / multiome)

```{r tenx, eval = FALSE}
# BiocManager::install("DropletUtils")
library(DropletUtils); library(SingleCellExperiment)
sce <- read10xCounts("filtered_feature_bc_matrix.h5", col.names = TRUE)
table(rowData(sce)$Type)                          # "Gene Expression"/"Antibody Capture"/"Peaks"

sce <- splitAltExps(sce, rowData(sce)$Type, ref = "Gene Expression")  # RNA -> main
altExpNames(sce)[altExpNames(sce) == "Antibody Capture"] <- "ADT"
assayNames(sce)[1] <- "counts"
sce$cell_type <- my_labels                        # supply your annotation
```

⚠️ **10x multiome ATAC is *peaks*, but Matilda needs a *gene-activity score*.**
Convert peaks → gene activity first (Signac), then attach:

```{r geneact, eval = FALSE}
# library(Signac); library(Seurat)
ga <- GeneActivity(seurat_atac_object)            # peaks -> gene-level activity
altExp(sce, "ATAC") <- SingleCellExperiment(list(counts = ga[, colnames(sce)]))
```

## From a Seurat object

```{r seurat, eval = FALSE}
library(Seurat); library(SingleCellExperiment)
sce <- as.SingleCellExperiment(seurat_obj)        # active assay -> main
# attach the other modalities (raw counts; ATAC as gene activity):
altExp(sce, "ADT")  <- SingleCellExperiment(list(counts = GetAssayData(seurat_obj, "ADT",  "counts")))
altExp(sce, "ATAC") <- SingleCellExperiment(list(counts = GetAssayData(seurat_obj, "ATAC", "counts")))
sce$cell_type <- seurat_obj$cell_type
```

(Matilda also takes a `Seurat` object directly once ADT/ATAC are altExps.)

## From plain matrices (no SCE needed)

The simplest path — pass genes × cells matrices and a label vector aligned to columns:

```{r matrices, eval = FALSE}
fit <- matilda_train(rna = rna_mat, adt = adt_mat, atac = atac_mat, cty = labels)
```

## From Matilda's own `.h5` (`matrix/data`, `matrix/features`, `matrix/barcodes`)

```{r matildah5, eval = FALSE}
source("_setup.R")                                # provides read_matilda_h5()
rna_mat <- read_matilda_h5("train_rna.h5")        # -> genes × cells matrix
```

## Preprocessing checklist

1. **Counts, not normalised** — RNA/ADT raw counts; ATAC gene-activity score.
2. **QC filter** — drop empty/low-count cells and genes detected in very few cells
   (e.g. < 1% of cells) in each modality.
3. **One cell set** — identical cells, same column order, across all modalities.
4. **Labels** — a `colData` cell-type column (needed for training, markers, simulation).
5. **Train vs query features must match** — the query must use the training feature
   set. Either intersect to shared features, or align the query to the training
   features and zero-fill the missing ones:

```{r align, eval = FALSE}
aligned <- matrix(0, nrow(train_rna), ncol(query_rna),
                  dimnames = list(rownames(train_rna), colnames(query_rna)))
shared <- intersect(rownames(train_rna), rownames(query_rna))
aligned[shared, ] <- query_rna[shared, ]
query_rna <- aligned
```

## Sanity-check before training

```{r check}
check_matilda_input <- function(sce, label = "cell_type") {
  stopifnot(methods::is(sce, "SingleCellExperiment"))
  stopifnot(label %in% colnames(SummarizedExperiment::colData(sce)))
  aes <- SingleCellExperiment::altExps(sce)
  stopifnot(all(vapply(aes, function(a) ncol(a) == ncol(sce), logical(1))))
  message("OK: ", ncol(sce), " cells, RNA ", nrow(sce), " features; altExps: ",
          paste(SingleCellExperiment::altExpNames(sce), collapse = ", "))
  invisible(TRUE)
}
check_matilda_input(matilda_example_sce())
```

Once this passes, continue to [training](01-training.Rmd).
