---
title: "Matilda tutorial 1 — Training"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Matilda trains a single multimodal variational autoencoder + classifier, then
reuses it for every downstream task. This tutorial trains it on the TEA-seq demo
and caches the result for tutorials 2–5.

> **Requirements:** the TEA-seq demo (`matilda_example_teaseq()`), and the bundled
> Python environment (built automatically on first use; GPU used if present).

## Load the data as a SingleCellExperiment

```{r load}
source("_setup.R")                      # read_matilda_h5(), load_teaseq_sce()
ref <- load_teaseq_sce("train")         # RNA in the main assay; ADT/ATAC as altExps
ref
```

## Train

```{r train}
ref <- matilda_train(ref, label = "cell_type")   # ~12 s on a GPU
matilda_model(ref)                                # the trained model lives in the object
```

The model is stored in `metadata(ref)$matilda`, so `ref` pipes directly into the
task tutorials. Cache it for reuse:

```{r save}
saveRDS(ref, "matilda_teaseq_ref.rds")
```

Next: [classification](02-classification.Rmd).
