---
title: "Matilda tutorial 2 — Classification"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Predict cell types for **query** cells by projecting them onto the trained
reference latent space.

```{r setup}
source("_setup.R")
ref <- if (file.exists("matilda_teaseq_ref.rds")) readRDS("matilda_teaseq_ref.rds") else
  matilda_train(load_teaseq_sce("train"), label = "cell_type")   # run tutorial 1 to cache
query <- load_teaseq_sce("test")
```

## Classify

```{r classify}
query <- matilda_classify(query, reference = ref)
head(data.frame(predicted = query$matilda_pred,
                probability = round(query$matilda_prob, 2)))
```

```{r accuracy}
acc <- mean(query$matilda_pred == query$cell_type)
sprintf("query accuracy = %.4f", acc)
```

## Performance validation

On TEA-seq this gives **accuracy 0.8092**, **identical** to the original Matilda
Python implementation run in the same environment (100 % per-cell label
agreement). Reproduce the head-to-head check with `inst/scripts/parity_check.R`.

Next: [dimension reduction](03-dimension-reduction.Rmd).
