---
title: "Matilda tutorial 4 — Feature selection"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Rank cell-type marker features across **all modalities** at once, attributing the
classifier's predictions back to the inputs with integrated gradients.

```{r setup}
source("_setup.R")
ref <- if (file.exists("matilda_teaseq_ref.rds")) readRDS("matilda_teaseq_ref.rds") else
  matilda_train(load_teaseq_sce("train"), label = "cell_type")
```

## Per-cell-type markers

```{r markers}
markers <- matilda_markers(ref, method = "IntegratedGradient")
head(markers)                                   # celltype / feature / importance
```

```{r top}
# top 5 features for one cell type
head(markers[markers$celltype == "B.Naive", ], 5)
```

`method = "Saliency"` is the faster gradient-magnitude alternative.

## Performance validation

Importance scores are **bit-identical** to the original Matilda Python
implementation (max|Δ| = 0 across all 11 cell types). See
`inst/scripts/parity_check.R`.

Next: [data simulation](05-simulation.Rmd).
