---
title: "Matilda on TEA-seq — R tutorial"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
library(ggplot2)
```

The complete Matilda workflow in **R** on real **TEA-seq** (RNA + ADT + ATAC): read your data →
train → classification → dimension reduction → feature selection → simulation, plus a look at the
different data types Matilda supports. (Python users: a parallel tutorial with the same structure
is available as a Jupyter notebook.)

## Requirements

- **R**: `matilda`, `SingleCellExperiment`; `Seurat` for the `10x` and `.rds` loaders; `scater`
  for UMAP; `ggplot2` for plots.
- **Python**: none — `matilda` provisions it via `basilisk` on the first `matilda_train()`.
- **Demo data**: downloaded automatically by `matilda_download_example()` (~75 MB, cached locally).

```{r deps}
library(matilda)
library(SingleCellExperiment)
data_dir <- matilda_download_example()         # downloads + caches the TEA-seq demo, returns the path
demo_h5  <- data_dir                            # native .h5 + _cty.csv
demo_fmt <- file.path(data_dir, "formats")      # .h5ad / 10x / .rds

# colour palettes used throughout
CELLTYPE_COLORS <- c(
  "B.Activated"="#1f77b4","B.Naive"="#aec7e8","DC.Myeloid"="#ff7f0e","Mono.CD14"="#ffbb78",
  "Mono.CD16"="#2ca02c","NK"="#98df8a","Platelets"="#d62728","T.CD4.Memory"="#ff9896",
  "T.CD4.Naive"="#9467bd","T.CD8.Effector"="#c5b0d5","T.CD8.Naive"="#8c564b")
ANCHOR_COLORS <- c("anchor cell type"="#F8766D","other cell types"="#00BFC4")
```

# 1. Read your data

Matilda's R interface takes a `SingleCellExperiment`: **RNA raw counts** in the main assay, **ADT**
and **ATAC** (gene activity) as `altExp`s, and a cell-type column in `colData`. Below the *same*
TEA-seq training cells are loaded from **four formats** — pick whichever matches your data.

```{r build-sce}
# assemble an SCE from three genes x cells matrices + a label vector
build_sce <- function(rna, adt, atac, cell_type) {
  sce <- SingleCellExperiment(assays = list(counts = as.matrix(rna)))
  altExp(sce, "ADT")  <- SummarizedExperiment(list(counts = as.matrix(adt)))
  altExp(sce, "ATAC") <- SummarizedExperiment(list(counts = as.matrix(atac)))
  sce$cell_type <- cell_type
  sce
}
labels <- as.character(utils::read.csv(file.path(demo_h5, "train_cty.csv"), header = FALSE)[[2]])[-1]
```

## Format A — native Matilda `.h5` (`rhdf5`)

```{r read-h5}
read_h5 <- function(path) {
  m     <- rhdf5::h5read(path, "matrix/data")
  feats <- as.character(rhdf5::h5read(path, "matrix/features"))
  cells <- as.character(rhdf5::h5read(path, "matrix/barcodes"))
  if (nrow(m) == length(cells) && ncol(m) == length(feats)) m <- t(m)   # -> genes x cells
  dimnames(m) <- list(feats, cells); m
}
g  <- function(mod) file.path(demo_h5, sprintf("train_%s.h5", mod))
sce <- build_sce(read_h5(g("rna")), read_h5(g("adt")), read_h5(g("atac")), labels)
sce
```

## Format B — `.h5ad` (AnnData / scanpy)

A `.h5ad` is just HDF5; read its CSR-sparse `X` with `rhdf5` (no extra Python needed). For a
richer object you could use `zellkonverter::readH5AD()` instead.

```{r read-h5ad}
rd_h5ad <- function(p) {                                      # -> genes x cells
  X     <- rhdf5::h5read(p, "X")                              # CSR: data / indices / indptr
  shp   <- as.integer(rhdf5::h5readAttributes(p, "X")$shape)  # c(cells, genes)
  genes <- as.character(rhdf5::h5read(p, "var/_index"))
  cells <- as.character(rhdf5::h5read(p, "obs/_index"))
  m <- Matrix::sparseMatrix(i = rep(seq_len(shp[1]), diff(X$indptr)),
                            j = X$indices + 1L, x = as.numeric(X$data), dims = shp)  # cells x genes
  m <- Matrix::t(m); dimnames(m) <- list(genes, cells); as.matrix(m)
}
h <- function(mod) file.path(demo_fmt, sprintf("train_%s.h5ad", mod))
sce_h5ad <- build_sce(rd_h5ad(h("rna")), rd_h5ad(h("adt")), rd_h5ad(h("atac")), labels)
```

## Format C — 10x CellRanger (`Seurat::Read10X`)

```{r read-10x, eval = requireNamespace("Seurat", quietly = TRUE)}
rd_10x <- function(d) as.matrix(Seurat::Read10X(d))
x <- function(mod) file.path(demo_fmt, "10x", sprintf("train_%s", mod))
sce_10x <- build_sce(rd_10x(x("rna")), rd_10x(x("adt")), rd_10x(x("atac")), labels)
```

## Format D — a `Seurat` object (`.rds`)

```{r read-rds, eval = requireNamespace("Seurat", quietly = TRUE)}
seu  <- readRDS(file.path(demo_fmt, "train_seurat.rds"))            # RNA/ADT/ATAC assays + cell_type
ga   <- function(a) as.matrix(SeuratObject::LayerData(seu, assay = a, layer = "counts"))
sce_rds <- build_sce(ga("RNA"), ga("ADT"), ga("ATAC"), seu$cell_type)
```

All four give the same object, so the rest of the tutorial doesn't depend on the format. We
continue with `sce` from Format A.

# 2. Train the Matilda model

The model is stored inside the object, so the workflow pipes.

```{r train}
sce <- matilda_train(sce, label = "cell_type", seed = 1L)
matilda_model(sce)
```

# 3. Classification (held-out query cells)

```{r classify}
query <- build_sce(read_h5(file.path(demo_h5, "test_rna.h5")),
                   read_h5(file.path(demo_h5, "test_adt.h5")),
                   read_h5(file.path(demo_h5, "test_atac.h5")),
                   as.character(utils::read.csv(file.path(demo_h5, "test_cty.csv"), header = FALSE)[[2]])[-1])
query <- matilda_classify(query, reference = sce)
sprintf("overall query accuracy = %.4f", mean(query$matilda_pred == query$cell_type))
```

```{r classify-plot, fig.width = 7, fig.height = 5}
acc <- tapply(query$matilda_pred == query$cell_type, query$cell_type, mean)
ggplot(data.frame(cell_type = names(acc), accuracy = as.numeric(acc)),
       aes(reorder(cell_type, accuracy), accuracy)) +
  geom_col(fill = "steelblue") + coord_flip() + ylim(0, 1) +
  geom_text(aes(label = sprintf("%.2f", accuracy)), hjust = -0.1, size = 3) +
  labs(title = "Query accuracy by cell type", x = NULL, y = "accuracy") + theme_classic()
```

# 4. Dimension reduction

```{r reduce}
sce <- matilda_reduce(sce)                 # adds reducedDim(sce, "MATILDA")
dim(reducedDim(sce, "MATILDA"))            # cells x z_dim (100)
```

```{r umap, fig.width = 7, fig.height = 5, eval = requireNamespace("scater", quietly = TRUE)}
set.seed(42)
sce <- scater::runUMAP(sce, dimred = "MATILDA", n_neighbors = 15L, min_dist = 0.1)
scater::plotUMAP(sce, colour_by = "cell_type") +
  ggplot2::scale_colour_manual(values = CELLTYPE_COLORS) +
  ggtitle("Matilda integrated latent space (UMAP)")
```

# 5. Feature selection

```{r markers}
markers <- matilda_markers(sce)            # data.frame: celltype / feature / importance
head(markers)
```

```{r markers-plot, fig.width = 8, fig.height = 6}
# order cell types alphabetically; group features by the cell type they top-3 mark, then importance
ord_ct <- sort(unique(markers$celltype))
top    <- do.call(rbind, lapply(ord_ct, function(ct) {
  d <- markers[markers$celltype == ct, ]; head(d[order(-d$importance), ], 3)
}))
feat_order <- unique(top$feature)
sub <- markers[markers$feature %in% feat_order, ]
sub$imp_s    <- ave(sub$importance, sub$feature, FUN = function(z) (z - min(z)) / (max(z) - min(z) + 1e-9))
sub$feature  <- factor(sub$feature,  levels = rev(feat_order))
sub$celltype <- factor(sub$celltype, levels = ord_ct)
ggplot(sub, aes(celltype, feature, fill = imp_s)) +
  geom_tile() + scale_fill_viridis_c() + theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Top-3 markers per cell type", x = NULL, y = NULL, fill = "scaled\nimportance")
```

# 6. Simulation

Matilda augments a chosen **anchor** cell type and returns the synthetic cells. The returned object
also carries the real reference cells Matilda used (`metadata(sim)$real`), in the **same feature
space** — so the two are directly comparable. The official check: per modality, UMAP the real and
simulated cells together, then show the real cells (top row) and the simulated cells (bottom row),
highlighting the anchor. If augmentation works, the anchor lands in the same place in both rows.

```{r simulate}
sim  <- matilda_simulate(sce, celltype = "B.Naive", n = 200)
real <- S4Vectors::metadata(sim)$real        # real reference cells, same space as the simulation
c(simulated = ncol(sim), real = ncol(real$rna), anchor_cells = sum(sim$label == "B.Naive"))
```

```{r sim-plot, fig.width = 9, fig.height = 6.5, eval = requireNamespace("scater", quietly = TRUE)}
real_mat <- list(RNA = real$rna, ADT = real$adt, ATAC = real$atac)
sim_mat  <- list(RNA  = SummarizedExperiment::assay(sim, "counts"),
                 ADT  = SummarizedExperiment::assay(SingleCellExperiment::altExp(sim, "ADT"),  "counts"),
                 ATAC = SummarizedExperiment::assay(SingleCellExperiment::altExp(sim, "ATAC"), "counts"))
anchor_tag <- function(lab) factor(ifelse(lab == "B.Naive", "anchor cell type", "other cell types"),
                                   levels = c("anchor cell type", "other cell types"))
one_modality <- function(mod) {                # UMAP real+simulated together, then split the layout
  comb <- cbind(real_mat[[mod]], sim_mat[[mod]])
  nr <- ncol(real_mat[[mod]]); ns <- ncol(sim_mat[[mod]])
  sce_c <- SingleCellExperiment::SingleCellExperiment(list(logcounts = comb))
  set.seed(42)
  sce_c <- scater::runUMAP(sce_c, n_neighbors = 20L, scale = TRUE, ntop = 1000)
  u <- as.data.frame(SingleCellExperiment::reducedDim(sce_c, "UMAP")); names(u) <- c("UMAP1", "UMAP2")
  rbind(data.frame(u[seq_len(nr), ],      grp = anchor_tag(real$label), panel = paste("Real", mod)),
        data.frame(u[nr + seq_len(ns), ], grp = anchor_tag(sim$label),  panel = paste("Simulated", mod)))
}
big <- do.call(rbind, lapply(c("RNA", "ADT", "ATAC"), one_modality))
big$panel <- factor(big$panel, levels = c("Real RNA", "Real ADT", "Real ATAC",
                                          "Simulated RNA", "Simulated ADT", "Simulated ATAC"))
ggplot(big, aes(UMAP1, UMAP2, colour = grp)) +
  geom_point(size = 0.5, alpha = 0.8) +
  facet_wrap(~ panel, nrow = 2, scales = "free") +
  ggplot2::scale_colour_manual(values = ANCHOR_COLORS) +
  theme_classic() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        legend.title = element_blank(), aspect.ratio = 1) +
  labs(x = NULL, y = NULL, title = "Real vs. simulated cells per modality (anchor: B.Naive)")
```

# 7. The data types Matilda supports

Matilda auto-detects the **mode** from the modalities present — `rna_only`, `CITEseq` (RNA+ADT),
`SHAREseq` (RNA+ATAC), `TEAseq` (RNA+ADT+ATAC). Training each on the *same* real TEA-seq cells
(a modality ablation — only TEA-seq ships with Matilda) shows what each modality buys; to run your
own data of a type, just build the SCE with those modalities.

```{r modality, fig.width = 7, fig.height = 4}
keep_only <- function(s, keep) { for (ae in altExpNames(s)) if (!(ae %in% keep)) altExp(s, ae) <- NULL; s }
acc <- sapply(list("RNA only" = character(0), "RNA+ATAC" = "ATAC", "RNA+ADT" = "ADT"),
  function(keep) {
    fit <- matilda_train(keep_only(sce, keep), label = "cell_type", seed = 1L)
    qy  <- matilda_classify(keep_only(query, keep), reference = fit)
    mean(qy$matilda_pred == qy$cell_type)
  })
acc["RNA+ADT+ATAC"] <- mean(query$matilda_pred == query$cell_type)   # the TEAseq model from §2/§3
ggplot(data.frame(modalities = factor(names(acc), levels = names(acc)), accuracy = as.numeric(acc)),
       aes(modalities, accuracy)) +
  geom_col(fill = "seagreen") + ylim(0, 1) +
  geom_text(aes(label = sprintf("%.3f", accuracy)), vjust = -0.4, size = 3.5) +
  labs(title = "Query accuracy by modality combination", x = NULL, y = "accuracy") + theme_classic()
```

Adding the ADT (protein) panel to RNA gives the biggest gain; ATAC contributes less for this task
on this dataset.

# 8. Session info

```{r sessioninfo}
sessionInfo()
```
