---
title: "Matilda on TEA-seq — R tutorial"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
library(ggplot2)
```

The complete Matilda workflow in **R** on real **TEA-seq** (RNA + ADT + ATAC): read your data →
train → classification → dimension reduction → feature selection → simulation, plus a look at the
different data types Matilda supports. (A parallel **Python** tutorial, `matilda-tutorial.ipynb`,
has the identical structure.)

## Requirements

- **R**: `matilda`, `SingleCellExperiment`; for the alternative loaders `zellkonverter` (`.h5ad`)
  and `Seurat` (`10x` + `.rds`); `scater` for UMAP; `ggplot2` for plots.
- **Python**: none — `matilda` provisions it via `basilisk` on the first `matilda_train()`.
- **Demo data**: TEA-seq in four formats (paths below).

```{r deps}
library(matilda)
library(SingleCellExperiment)
demo_h5  <- getOption("matilda.demo",         "~/Documents/matilda-r/teaseq")           # native .h5 + _cty.csv
demo_fmt <- getOption("matilda.demo_formats", "~/Documents/matilda-r/teaseq_formats")   # .h5ad / 10x / .rds
```

# 1. Read your data

Matilda's R interface takes a `SingleCellExperiment`: **RNA raw counts** in the main assay, **ADT**
and **ATAC** (gene activity) as `altExp`s, and a cell-type column in `colData`. Below the *same*
TEA-seq training cells are loaded from **four formats** — pick whichever matches your data.

```{r build-sce}
# assemble an SCE from three genes x cells matrices + a label vector
build_sce <- function(rna, adt, atac, cell_type) {
  sce <- SingleCellExperiment(assays = list(counts = as.matrix(rna)))
  altExp(sce, "ADT")  <- SummarizedExperiment(list(counts = as.matrix(adt)))
  altExp(sce, "ATAC") <- SummarizedExperiment(list(counts = as.matrix(atac)))
  sce$cell_type <- cell_type
  sce
}
labels <- as.character(utils::read.csv(file.path(demo_h5, "train_cty.csv"), header = FALSE)[[2]])[-1]
```

## Format A — native Matilda `.h5` (`rhdf5`)

```{r read-h5}
read_h5 <- function(path) {
  m     <- rhdf5::h5read(path, "matrix/data")
  feats <- as.character(rhdf5::h5read(path, "matrix/features"))
  cells <- as.character(rhdf5::h5read(path, "matrix/barcodes"))
  if (nrow(m) == length(cells) && ncol(m) == length(feats)) m <- t(m)   # -> genes x cells
  dimnames(m) <- list(feats, cells); m
}
g  <- function(mod) file.path(demo_h5, sprintf("train_%s.h5", mod))
sce <- build_sce(read_h5(g("rna")), read_h5(g("adt")), read_h5(g("atac")), labels)
sce
```

## Format B — `.h5ad` (AnnData / scanpy, via `zellkonverter`)

```{r read-h5ad, eval = requireNamespace("zellkonverter", quietly = TRUE)}
rd_h5ad <- function(p) as.matrix(SummarizedExperiment::assay(zellkonverter::readH5AD(p), "counts"))
h <- function(mod) file.path(demo_fmt, sprintf("train_%s.h5ad", mod))
sce_h5ad <- build_sce(rd_h5ad(h("rna")), rd_h5ad(h("adt")), rd_h5ad(h("atac")), labels)
isTRUE(all.equal(as.matrix(counts(sce)), as.matrix(counts(sce_h5ad))))   # identical to Format A
```

## Format C — 10x CellRanger (`Seurat::Read10X`)

```{r read-10x, eval = requireNamespace("Seurat", quietly = TRUE)}
rd_10x <- function(d) as.matrix(Seurat::Read10X(d))
x <- function(mod) file.path(demo_fmt, "10x", sprintf("train_%s", mod))
sce_10x <- build_sce(rd_10x(x("rna")), rd_10x(x("adt")), rd_10x(x("atac")), labels)
isTRUE(all.equal(as.matrix(counts(sce)), as.matrix(counts(sce_10x))))    # identical to Format A
```

## Format D — a `Seurat` object (`.rds`)

```{r read-rds, eval = requireNamespace("Seurat", quietly = TRUE)}
seu  <- readRDS(file.path(demo_fmt, "train_seurat.rds"))            # RNA/ADT/ATAC assays + cell_type
ga   <- function(a) as.matrix(SeuratObject::LayerData(seu, assay = a, layer = "counts"))
sce_rds <- build_sce(ga("RNA"), ga("ADT"), ga("ATAC"), seu$cell_type)
isTRUE(all.equal(as.matrix(counts(sce)), as.matrix(counts(sce_rds))))    # identical to Format A
```

All four give the same object, so the rest of the tutorial doesn't depend on the format. We
continue with `sce` from Format A.

# 2. Train the Matilda model

The model is stored inside the object, so the workflow pipes.

```{r train}
sce <- matilda_train(sce, label = "cell_type")
matilda_model(sce)
```

# 3. Classification (held-out query cells)

```{r classify}
query <- build_sce(read_h5(file.path(demo_h5, "test_rna.h5")),
                   read_h5(file.path(demo_h5, "test_adt.h5")),
                   read_h5(file.path(demo_h5, "test_atac.h5")),
                   as.character(utils::read.csv(file.path(demo_h5, "test_cty.csv"), header = FALSE)[[2]])[-1])
query <- matilda_classify(query, reference = sce)
sprintf("overall query accuracy = %.4f", mean(query$matilda_pred == query$cell_type))
```

```{r classify-plot, fig.width = 7, fig.height = 5}
acc <- tapply(query$matilda_pred == query$cell_type, query$cell_type, mean)
ggplot(data.frame(cell_type = names(acc), accuracy = as.numeric(acc)),
       aes(reorder(cell_type, accuracy), accuracy)) +
  geom_col(fill = "steelblue") + coord_flip() + ylim(0, 1) +
  geom_text(aes(label = sprintf("%.2f", accuracy)), hjust = -0.1, size = 3) +
  labs(title = "Query accuracy by cell type", x = NULL, y = "accuracy") + theme_classic()
```

# 4. Dimension reduction

```{r reduce}
sce <- matilda_reduce(sce)                 # adds reducedDim(sce, "MATILDA")
dim(reducedDim(sce, "MATILDA"))            # cells x z_dim (100)
```

```{r umap, fig.width = 7, fig.height = 5, eval = requireNamespace("scater", quietly = TRUE)}
sce <- scater::runUMAP(sce, dimred = "MATILDA")
scater::plotUMAP(sce, colour_by = "cell_type") + ggtitle("Matilda integrated latent space (UMAP)")
```

# 5. Feature selection

```{r markers}
markers <- matilda_markers(sce)            # data.frame: celltype / feature / importance
head(markers)
```

```{r markers-plot, fig.width = 8, fig.height = 6}
top <- do.call(rbind, by(markers, markers$celltype,
                         function(d) head(d[order(-d$importance), ], 3)))
sub <- markers[markers$feature %in% unique(top$feature), ]
sub$imp_s <- ave(sub$importance, sub$feature, FUN = function(z) (z - min(z)) / (max(z) - min(z) + 1e-9))
ggplot(sub, aes(celltype, feature, fill = imp_s)) +
  geom_tile() + scale_fill_viridis_c() + theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Top-3 markers per cell type", x = NULL, y = NULL, fill = "scaled\nimportance")
```

# 6. Simulation

```{r simulate}
sim <- matilda_simulate(sce, celltype = "B.Naive", n = 200)
sim$origin <- ifelse(sim$label == "B.Naive", "synthetic B.Naive", "real (other types)")
table(sim$origin)
```

```{r sim-plot, fig.width = 6, fig.height = 5, eval = requireNamespace("scater", quietly = TRUE)}
sim <- scater::runUMAP(sim, exprs_values = "counts", scale = TRUE, ntop = 1000, n_neighbors = 20)
scater::plotUMAP(sim, colour_by = "origin") + ggtitle("Simulated B.Naive cells among the real cells")
```

# 7. The data types Matilda supports

Matilda auto-detects the **mode** from the modalities present — `rna_only`, `CITEseq` (RNA+ADT),
`SHAREseq` (RNA+ATAC), `TEAseq` (RNA+ADT+ATAC). Training each on the *same* real TEA-seq cells
(a modality ablation — only TEA-seq ships with Matilda) shows what each modality buys; to run your
own data of a type, just build the SCE with those modalities.

```{r modality, fig.width = 7, fig.height = 4}
keep_only <- function(s, keep) { for (ae in altExpNames(s)) if (!(ae %in% keep)) altExp(s, ae) <- NULL; s }
acc <- sapply(list("RNA only" = character(0), "RNA+ATAC" = "ATAC", "RNA+ADT" = "ADT"),
  function(keep) {
    fit <- matilda_train(keep_only(sce, keep), label = "cell_type", seed = 1L)
    qy  <- matilda_classify(keep_only(query, keep), reference = fit)
    mean(qy$matilda_pred == qy$cell_type)
  })
acc["RNA+ADT+ATAC"] <- mean(query$matilda_pred == query$cell_type)   # the TEAseq model from §2/§3
ggplot(data.frame(modalities = factor(names(acc), levels = names(acc)), accuracy = as.numeric(acc)),
       aes(modalities, accuracy)) +
  geom_col(fill = "seagreen") + ylim(0, 1) +
  geom_text(aes(label = sprintf("%.3f", accuracy)), vjust = -0.4, size = 3.5) +
  labs(title = "Query accuracy by modality combination", x = NULL, y = "accuracy") + theme_classic()
```

Adding the ADT (protein) panel to RNA gives the biggest gain; ATAC contributes less for this task
on this dataset.

# 8. Reproducibility

The R wrapper runs Matilda's **unchanged** Python, so on the **same hardware** it is bit-identical
to the original (verified across all tasks by `inst/scripts/parity_check.R`, `max|Δ| = 0`):
GPU query accuracy 0.8092, CPU 0.8148. The accuracy in §3 is this machine's run; the small GPU↔CPU
gap is PyTorch floating-point, not the wrapper. Results are format-independent (the §1 checks are
all `TRUE`).

```{r sessioninfo}
sessionInfo()
```
