---
title: "Matilda on TEA-seq: a Python ⇄ R mirror (data → tasks → plots)"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
library(ggplot2)
```

This tutorial is a **one-to-one mirror of the original Matilda Python tutorial**, on the
real **TEA-seq** dataset (RNA + ADT + ATAC). For every step it shows the **Python** command
you would run and the **R** call that runs the *same* unchanged code underneath, then the
output and a plot. The R wrapper drives Matilda's vendored Python via `basilisk`, so on the
same hardware the two are **bit-identical** (verified across all tasks by
`inst/scripts/parity_check.R`).

Layout mirrors the docs: **train → multi-task on the training data → multi-task on the query
data**. Plots mirror the official `qc/visulize_latent_space.Rmd` and
`qc/visualize_simulated_data.Rmd`.

## Requirements

- **R**: `matilda`, `SingleCellExperiment`; `Seurat` + `zellkonverter` for the alternative
  loaders; `scater` for UMAPs; `ggplot2` for the plots (chunks skip if a package is missing).
- **Python**: none — provisioned by `basilisk` on the first `matilda_train()` (one-off build).
- **Demo data**: the 8 TEA-seq files (`{train,test}_{rna,adt,atac}.h5` + `{train,test}_cty.csv`).

```{r deps}
library(matilda)
library(SingleCellExperiment)
demo_h5  <- getOption("matilda.demo",         "~/Documents/matilda-r/teaseq")
demo_fmt <- getOption("matilda.demo_formats", "~/Documents/matilda-r/teaseq_formats")
```

# 1. Read the data

Matilda's input is **RNA raw counts** + **ADT** + **ATAC** gene-activity. Python reads the
`.h5`/`.csv` files directly inside the CLI (`util.read_h5_data`); in R we read the same files
into a `SingleCellExperiment` (RNA in the main assay, ADT/ATAC as `altExp`s, labels in
`colData`). The reading is explicit so you can adapt it to your own data.

**Python** — the files are passed straight to the scripts, e.g. `--rna train_rna.h5 ...`.

**R** — read the native Matilda `.h5` (`matrix/data` / `features` / `barcodes`):

```{r load-h5}
read_h5 <- function(path) {
  m     <- rhdf5::h5read(path, "matrix/data")
  feats <- as.character(rhdf5::h5read(path, "matrix/features"))
  cells <- as.character(rhdf5::h5read(path, "matrix/barcodes"))
  if (nrow(m) == length(cells) && ncol(m) == length(feats)) m <- t(m)  # -> genes x cells
  stopifnot(nrow(m) == length(feats), ncol(m) == length(cells))
  dimnames(m) <- list(feats, cells); m
}
load_h5_sce <- function(dir, split) {
  g <- function(mod) file.path(dir, sprintf("%s_%s.h5", split, mod))
  sce <- SingleCellExperiment(assays = list(counts = read_h5(g("rna"))))
  altExp(sce, "ADT")  <- SummarizedExperiment(list(counts = read_h5(g("adt"))))
  altExp(sce, "ATAC") <- SummarizedExperiment(list(counts = read_h5(g("atac"))))
  sce$cell_type <- as.character(
    utils::read.csv(file.path(dir, sprintf("%s_cty.csv", split)), header = FALSE)[[2]])[-1]
  sce
}
sce <- load_h5_sce(demo_h5, "train")
sce
lapply(altExps(sce), dim)
```

The same data also loads from a **Seurat** object or **`.h5ad`** (AnnData), verified
element-identical to the `.h5` above — so downstream results don't depend on the format:

```{r load-seurat, eval = requireNamespace("Seurat", quietly = TRUE)}
seu <- readRDS(file.path(demo_fmt, "train_seurat.rds"))
ga  <- function(a) as.matrix(SeuratObject::LayerData(seu, assay = a, layer = "counts"))
c(RNA  = isTRUE(all.equal(as.matrix(counts(sce)),               ga("RNA"))),
  ADT  = isTRUE(all.equal(as.matrix(assay(altExp(sce, "ADT"))),  ga("ADT"))),
  ATAC = isTRUE(all.equal(as.matrix(assay(altExp(sce, "ATAC"))), ga("ATAC"))))
```

```{r load-h5ad, eval = requireNamespace("zellkonverter", quietly = TRUE)}
ad <- zellkonverter::readH5AD(file.path(demo_fmt, "train_rna.h5ad"))
isTRUE(all.equal(as.matrix(counts(sce)),
                 as.matrix(SummarizedExperiment::assay(ad, "counts"))))  # RNA identical
```

> Other starting points: 10x via `DropletUtils::read10xCounts()` (+ `splitAltExps()`); ATAC
> peaks must first become a gene-activity matrix (`Signac::GeneActivity()`).

# 2. Train the Matilda model

**Python**
```bash
python main_matilda_train.py \
  --rna train_rna.h5 --adt train_adt.h5 --atac train_atac.h5 --cty train_cty.csv \
  --batch_size 64 --epochs 30 --lr 0.02 --z_dim 100 \
  --hidden_rna 185 --hidden_adt 30 --hidden_atac 185 --seed 1
```

**R** (same defaults; the model is stored in the object and the workflow pipes):
```{r train}
sce <- matilda_train(sce, label = "cell_type")
matilda_model(sce)
```

# 3. Multi-task on the training data

## 3a. Simulation — `--simulation True --simulation_ct B.Naive --simulation_num 200`

**Python**
```bash
python main_matilda_task.py --rna train_rna.h5 --adt train_adt.h5 --atac train_atac.h5 \
  --cty train_cty.csv --simulation True --simulation_ct B.Naive --simulation_num 200
```

**R** — generate 200 synthetic `B.Naive` cells (mirrors `qc/visualize_simulated_data.Rmd`):
```{r sim-train}
sim <- matilda_simulate(sce, celltype = "B.Naive", n = 200)
sim$origin <- ifelse(sim$label == "B.Naive", "synthetic B.Naive", "real (other types)")
table(sim$origin)
```

```{r sim-plot, fig.width = 6, fig.height = 5, eval = requireNamespace("scater", quietly = TRUE)}
sim <- scater::runUMAP(sim, exprs_values = "counts", scale = TRUE, ntop = 1000, n_neighbors = 20)
scater::plotUMAP(sim, colour_by = "origin") +
  ggtitle("Simulated B.Naive cells among the real cells")
```

## 3b. Dimension reduction — `--dim_reduce True`

**Python**
```bash
python main_matilda_task.py --rna train_rna.h5 --adt train_adt.h5 --atac train_atac.h5 \
  --cty train_cty.csv --dim_reduce True
```

**R**
```{r reduce-train}
sce <- matilda_reduce(sce)
dim(reducedDim(sce, "MATILDA"))            # cells x z_dim (100)
```

Plot the latent space, reproducing `qc/visulize_latent_space.Rmd` (Seurat: PCA → UMAP on the
first 10 PCs):
```{r umap-train, fig.width = 7, fig.height = 5, eval = requireNamespace("Seurat", quietly = TRUE)}
L <- as(t(reducedDim(sce, "MATILDA")), "CsparseMatrix")
rownames(L) <- paste0("z", seq_len(nrow(L))); colnames(L) <- colnames(sce)
su <- Seurat::CreateSeuratObject(counts = L)
su[["RNA"]]$data <- su[["RNA"]]$counts
su$celltype <- sce$cell_type
su <- Seurat::RunPCA(Seurat::ScaleData(Seurat::FindVariableFeatures(su), verbose = FALSE), verbose = FALSE)
su <- Seurat::RunUMAP(su, dims = 1:10, verbose = FALSE)
Seurat::DimPlot(su, reduction = "umap", group.by = "celltype")
```

## 3c. Feature selection — `--fs True --fs_method IntegratedGradient`

**Python**
```bash
python main_matilda_task.py --rna train_rna.h5 --adt train_adt.h5 --atac train_atac.h5 \
  --cty train_cty.csv --fs True --fs_method IntegratedGradient
```

**R**
```{r markers}
markers <- matilda_markers(sce)            # data.frame: celltype / feature / importance
head(markers)
```

Top markers per cell type (importance scaled per feature for visibility):
```{r markers-plot, fig.width = 8, fig.height = 6}
top  <- do.call(rbind, by(markers, markers$celltype,
                          function(d) head(d[order(-d$importance), ], 3)))
sub  <- markers[markers$feature %in% unique(top$feature), ]
sub$imp_s <- ave(sub$importance, sub$feature,
                 FUN = function(z) (z - min(z)) / (max(z) - min(z) + 1e-9))
ggplot(sub, aes(celltype, feature, fill = imp_s)) +
  geom_tile() + scale_fill_viridis_c() + theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Top-3 markers per cell type", x = NULL, y = NULL, fill = "scaled\nimportance")
```

# 4. Multi-task on the query data

Same three task verbs, now on the held-out query split with `reference = sce` (the
`--query True` flag).

```{r load-query}
query <- load_h5_sce(demo_h5, "test")      # 5048 held-out cells
```

## 4a. Classification — `--classification True --query True`

**Python**
```bash
python main_matilda_task.py --rna test_rna.h5 --adt test_adt.h5 --atac test_atac.h5 \
  --cty test_cty.csv --classification True --query True
```

**R**
```{r classify-query}
query <- matilda_classify(query, reference = sce)
sprintf("overall query accuracy = %.4f", mean(query$matilda_pred == query$cell_type))
```

Per-cell-type accuracy:
```{r classify-plot, fig.width = 7, fig.height = 5}
acc <- tapply(query$matilda_pred == query$cell_type, query$cell_type, mean)
acc_df <- data.frame(cell_type = names(acc), accuracy = as.numeric(acc))
ggplot(acc_df, aes(reorder(cell_type, accuracy), accuracy)) +
  geom_col(fill = "steelblue") + coord_flip() + ylim(0, 1) +
  geom_text(aes(label = sprintf("%.2f", accuracy)), hjust = -0.1, size = 3) +
  labs(title = "Query classification accuracy by cell type", x = NULL, y = "accuracy") +
  theme_classic()
```

## 4b. Dimension reduction — `--dim_reduce True --query True`

**Python**
```bash
python main_matilda_task.py --rna test_rna.h5 --adt test_adt.h5 --atac test_atac.h5 \
  --cty test_cty.csv --dim_reduce True --query True
```

**R**
```{r reduce-query}
query <- matilda_reduce(query, reference = sce)
dim(reducedDim(query, "MATILDA"))          # 5048 x 100
```

## 4c. Feature selection — `--fs True --query True`

**Python**
```bash
python main_matilda_task.py --rna test_rna.h5 --adt test_adt.h5 --atac test_atac.h5 \
  --cty test_cty.csv --fs True --query True
```

**R**
```{r fs-query}
qmarkers <- matilda_markers(query, reference = sce)
head(qmarkers)
```

# 5. Running each data type (and why multimodal helps)

Matilda supports four **data types**, auto-detected from the modalities you provide, and adapts
the model to each. The four runs below are a **controlled ablation on the *same real* TEA-seq
cells** (modalities included/excluded) — not four separate datasets and not synthetic data.
Because Matilda treats a "data type" as simply *which modalities are present*, this is exactly how
you would run a real dataset of each type — just load the modalities you have:

| Data type | mode | R: build the object with | Python script |
|-----------|------|--------------------------|---------------|
| scRNA-seq | `rna_only` | RNA `counts` only (no `altExp`) | `main_matilda_rna_train.py` / `…_rna_task.py` |
| CITE-seq  | `CITEseq`  | RNA + `altExp "ADT"` | `main_matilda_train.py` / `main_matilda_task.py` |
| SHARE-seq | `SHAREseq` | RNA + `altExp "ATAC"` | `main_matilda_train.py` / `main_matilda_task.py` |
| TEA-seq   | `TEAseq`   | RNA + `altExp "ADT"` + `altExp "ATAC"` | `main_matilda_train.py` / `main_matilda_task.py` |

> **No CITE-seq/SHARE-seq demo dataset ships with Matilda** (only TEA-seq), so here we ablate
> TEA-seq to exercise every mode on real data. To run your *own* data of a given type, build the
> object with exactly the modalities above — `matilda_train()` selects the mode automatically.

Training each combination on the *same* cells also shows what each modality buys for classification.

**Python** — same scripts, just fewer modality flags (the mode follows automatically; RNA-only
uses the `*_rna_*` scripts):
```bash
python main_matilda_rna_train.py --rna train_rna.h5 --cty train_cty.csv                      # rna_only
python main_matilda_train.py --rna train_rna.h5 --adt  train_adt.h5  --cty train_cty.csv     # CITEseq
python main_matilda_train.py --rna train_rna.h5 --atac train_atac.h5 --cty train_cty.csv     # SHAREseq
# TEAseq (all three) was trained in section 2
```

**R** — drop `altExp`s to change the mode, then re-train and classify the query:
```{r modality, fig.width = 7, fig.height = 4}
keep_only <- function(s, keep) { for (ae in altExpNames(s)) if (!(ae %in% keep)) altExp(s, ae) <- NULL; s }
acc <- sapply(list("RNA only" = character(0), "RNA+ATAC" = "ATAC", "RNA+ADT" = "ADT"),
  function(keep) {
    fit <- matilda_train(keep_only(sce, keep), label = "cell_type", seed = 1L)
    qy  <- matilda_classify(keep_only(query, keep), reference = fit)
    mean(qy$matilda_pred == qy$cell_type)
  })
acc["RNA+ADT+ATAC"] <- mean(query$matilda_pred == query$cell_type)   # reuse the TEAseq model from §2/§4
mdf <- data.frame(modalities = factor(names(acc), levels = names(acc)), accuracy = as.numeric(acc))
ggplot(mdf, aes(modalities, accuracy)) + geom_col(fill = "seagreen") + ylim(0, 1) +
  geom_text(aes(label = sprintf("%.3f", accuracy)), vjust = -0.4, size = 3.5) +
  labs(title = "Query accuracy by modality combination", x = NULL, y = "accuracy") + theme_classic()
```

Adding the **ADT** (protein) panel to RNA gives the biggest jump in classification accuracy — the
core value of multimodal integration. **ATAC** contributes less *for this task on this dataset*
(full TEAseq ≈ RNA+ADT). Use all modalities for the integrated latent space and simulation; for
pure classification here, the protein panel carries most of the signal.

# 6. Reproducibility

The R wrapper runs Matilda's **unchanged** Python, so on the **same hardware** it is
bit-identical to the original (verified across classification, dimension reduction, feature
selection and simulation by `inst/scripts/parity_check.R`, `max|Δ| = 0`): GPU query accuracy
**0.8092**, CPU **0.8148**. The accuracy printed in §4 is **this machine's** run; the small
GPU↔CPU gap is PyTorch floating-point, not the wrapper. Results are format-independent
(§1 equivalence checks are all `TRUE`). UMAP layouts differ run-to-run by construction
(stochastic embedding), but the cluster structure is stable.

> **A note on augmentation (the 5th task).** `matilda_train(augmentation = TRUE)` (the default)
> VAE-resamples rare cell types to balance training — aimed at heavily imbalanced data. Its
> benefit is dataset-dependent: on this small TEA-seq split it does not improve accuracy
> (aug on = 0.8092 vs off = 0.8360), because the rarest types have only ~3 training cells. It is
> exposed as a toggle rather than demonstrated as a win here.

```{r sessioninfo}
sessionInfo()
```
