---
title: "Matilda: an end-to-end multimodal workflow (TEA-seq)"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

This single tutorial takes a real **TEA-seq** dataset (RNA + ADT + ATAC) from
**whatever format you have it in** — Matilda `.h5`, a `Seurat` object, `.h5ad`
(AnnData/scanpy), or plain matrices — into a `SingleCellExperiment`, and then runs
the complete Matilda workflow: **train → dimension reduction → classification →
feature selection → simulation**. Data loading is shown explicitly, because the
point is to show you *how to run your own data*.

## Requirements

- **R packages**: `matilda`, `SingleCellExperiment` (always); `Seurat` and
  `zellkonverter` only for the Seurat / `.h5ad` loading options (those chunks skip
  themselves if the package is missing).
- **Python**: nothing to install — `matilda` provisions a private environment via
  `basilisk` on the **first `matilda_train()` call** (a one-off multi-minute build,
  cached afterwards; a GPU is used if present, otherwise CPU).
- **Demo data** (set the two paths below):
  - `demo_h5/`  → `train_rna.h5 train_adt.h5 train_atac.h5 train_cty.csv` and the
    matching `test_*` (8 files);
  - `demo_fmt/` → `train_seurat.rds` and `train_{rna,adt,atac}.h5ad`
    (only needed for loading Options B / C).

```{r deps}
library(matilda)
library(SingleCellExperiment)
demo_h5  <- getOption("matilda.demo",         "~/Documents/matilda-r/teaseq")
demo_fmt <- getOption("matilda.demo_formats", "~/Documents/matilda-r/teaseq_formats")
```

# 1. What Matilda expects

A `SingleCellExperiment` with **RNA raw counts** in the main assay, **ADT** and
**ATAC** as `altExp`s, and a cell-type column in `colData`. ATAC must be a
gene-activity score (gene-level), not raw peaks. (Matilda also accepts a
`MultiAssayExperiment`, a `Seurat` object, or plain matrices.)

# 2. Load your data — pick the format you have

## Option A — native Matilda `.h5` (`matrix/data` / `features` / `barcodes`)

The reading is fully explicit (no hidden wrapper):

```{r load-h5}
read_h5 <- function(path) {
  m     <- rhdf5::h5read(path, "matrix/data")
  feats <- as.character(rhdf5::h5read(path, "matrix/features"))
  cells <- as.character(rhdf5::h5read(path, "matrix/barcodes"))
  if (nrow(m) == length(cells) && ncol(m) == length(feats)) m <- t(m)  # -> genes x cells
  stopifnot(nrow(m) == length(feats), ncol(m) == length(cells))
  dimnames(m) <- list(feats, cells)
  m
}
load_h5_sce <- function(dir, split) {
  g <- function(mod) file.path(dir, sprintf("%s_%s.h5", split, mod))
  sce <- SingleCellExperiment(assays = list(counts = read_h5(g("rna"))))
  altExp(sce, "ADT")  <- SummarizedExperiment(list(counts = read_h5(g("adt"))))
  altExp(sce, "ATAC") <- SummarizedExperiment(list(counts = read_h5(g("atac"))))
  sce$cell_type <- as.character(
    utils::read.csv(file.path(dir, sprintf("%s_cty.csv", split)), header = FALSE)[[2]])[-1]
  sce
}

sce <- load_h5_sce(demo_h5, "train")
sce
lapply(altExps(sce), dim)                 # ADT / ATAC: features x cells
```

This matches the section-1 contract: RNA counts in the main assay, ADT/ATAC as
`altExp`s, labels in `colData$cell_type`.

## Option B — a `Seurat` object

Needs `install.packages("Seurat")`; skip if you only have `.h5`.

```{r load-seurat, eval = requireNamespace("Seurat", quietly = TRUE)}
library(Seurat)
seu <- readRDS(file.path(demo_fmt, "train_seurat.rds"))   # RNA/ADT/ATAC assays + cell_type
ga  <- function(a) as.matrix(SeuratObject::LayerData(seu, assay = a, layer = "counts"))
sce_seurat <- SingleCellExperiment(assays = list(counts = ga("RNA")))
altExp(sce_seurat, "ADT")  <- SummarizedExperiment(list(counts = ga("ADT")))
altExp(sce_seurat, "ATAC") <- SummarizedExperiment(list(counts = ga("ATAC")))
sce_seurat$cell_type <- seu$cell_type

# identical to Option A — every modality + the labels:
c(RNA  = isTRUE(all.equal(as.matrix(counts(sce)),               ga("RNA"))),
  ADT  = isTRUE(all.equal(as.matrix(assay(altExp(sce, "ADT"))),  ga("ADT"))),
  ATAC = isTRUE(all.equal(as.matrix(assay(altExp(sce, "ATAC"))), ga("ATAC"))),
  labels = identical(as.character(sce$cell_type), as.character(sce_seurat$cell_type)))
```

## Option C — `.h5ad` (AnnData / scanpy), one file per modality

Needs `BiocManager::install("zellkonverter")`; skip if you only have `.h5`.

```{r load-h5ad, eval = requireNamespace("zellkonverter", quietly = TRUE)}
library(zellkonverter)
ad <- lapply(c(rna = "rna", adt = "adt", atac = "atac"),
             function(n) readH5AD(file.path(demo_fmt, sprintf("train_%s.h5ad", n))))
# select the counts assay by name (a user's AnnData may also carry logcounts):
am <- function(x) as.matrix(SummarizedExperiment::assay(x, "counts"))
sce_h5ad <- SingleCellExperiment(assays = list(counts = am(ad$rna)))
altExp(sce_h5ad, "ADT")  <- SummarizedExperiment(list(counts = am(ad$adt)))
altExp(sce_h5ad, "ATAC") <- SummarizedExperiment(list(counts = am(ad$atac)))
sce_h5ad$cell_type <- as.character(colData(ad$rna)$cell_type)

c(RNA  = isTRUE(all.equal(as.matrix(counts(sce)),               am(ad$rna))),
  ADT  = isTRUE(all.equal(as.matrix(assay(altExp(sce, "ADT"))),  am(ad$adt))),
  ATAC = isTRUE(all.equal(as.matrix(assay(altExp(sce, "ATAC"))), am(ad$atac))),
  labels = identical(as.character(sce$cell_type), as.character(sce_h5ad$cell_type)))
```

## Option D — plain matrices (no SingleCellExperiment)

```{r load-matrices, eval = FALSE}
# genes x cells matrices + a label vector; returns a model handle directly
fit <- matilda_train(rna = rna_mat, adt = adt_mat, atac = atac_mat, cty = labels)
```

Options **B and C are verified element-identical to Option A** above (every modality
and the labels return `TRUE`), so the downstream results don't depend on which
format you started from; Option D follows the same contract. We continue with `sce`
from Option A.

> **Other formats**: 10x CellRanger output reads via
> `DropletUtils::read10xCounts()` (split feature types into RNA/ADT with
> `splitAltExps()`); ATAC **peaks** must first be turned into a gene-activity
> matrix (e.g. `Signac::GeneActivity()`) before use.

# 3. Train, then run every task

The model is stored inside the object, so the workflow pipes:

```{r train}
sce <- matilda_train(sce, label = "cell_type")   # first call also builds the env (one-off)
matilda_model(sce)
```

```{r reduce}
sce <- matilda_reduce(sce)                        # adds reducedDim(sce, "MATILDA")
dim(reducedDim(sce, "MATILDA"))                   # cells x z_dim (100)
```

```{r reduce-plot, fig.width = 6, fig.height = 5}
# quick look at the integrated latent space (base graphics, no extra deps)
L <- reducedDim(sce, "MATILDA")
plot(L[, 1], L[, 2], col = factor(sce$cell_type), pch = 16, cex = 0.4,
     xlab = "latent 1", ylab = "latent 2", main = "Matilda integrated latent space")
# for a UMAP: sce <- scater::runUMAP(sce, dimred = "MATILDA"); scater::plotUMAP(sce, colour_by = "cell_type")
```

```{r classify}
sce <- matilda_classify(sce)                      # adds colData$matilda_pred / $matilda_prob
table(sce$matilda_pred)                           # predicted cell types
```

```{r markers}
markers <- matilda_markers(sce)                   # data.frame: celltype / feature / importance
head(markers)
```

```{r simulate}
# For a named cell type, matilda_simulate returns the reference cells with that
# type's cells replaced by `n` synthetic ones (a balanced reference); the synthetic
# cells carry the target label.
sim <- matilda_simulate(sce, celltype = "B.Naive", n = 200)
table(sim$label)                                  # B.Naive is now the 200 synthetic cells
sim[, sim$label == "B.Naive"]                     # extract just the synthetic cells
```

# 4. Classify held-out query cells

```{r query}
query <- load_h5_sce(demo_h5, "test")             # or load it from Seurat / .h5ad
query <- matilda_classify(query, reference = sce)
sprintf("query accuracy on this machine = %.4f", mean(query$matilda_pred == query$cell_type))
```

# 5. Reproducibility

The R wrapper runs Matilda's **unchanged** Python, so it reproduces the original
exactly **on the same hardware**. Verified bit-identical to the original Matilda
Python (same env + seed): on a **GPU**, query accuracy **0.8092** with per-cell
classification 100% concordant, and latent space / feature importance / simulation
all matching to `max|Δ| = 0`; on a **CPU**, R and Python agree exactly too
(both **0.8148**). Reproduce with `inst/scripts/parity_check.R`.

So the accuracy printed in section 4 is **this machine's run** (≈ 0.81 on a CPU
laptop); the reference GPU figure is 0.8092. The small gap is GPU-vs-CPU
floating-point — a property of PyTorch, not of this wrapper. Results are also
identical regardless of **input format**, since Options A–C load the same data
(the equivalence checks in section 2 are all `TRUE`).

```{r sessioninfo}
sessionInfo()
```
