---
title: "Matilda: an end-to-end multimodal workflow (TEA-seq)"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

This tutorial takes a real **TEA-seq** dataset (RNA + ADT + ATAC) from **whatever
format you have it in** — Matilda `.h5`, a `Seurat` object, `.h5ad` (AnnData/scanpy),
or plain matrices — into a `SingleCellExperiment`, and then runs the complete Matilda
workflow. The task sections mirror the official Python tutorial one-to-one:

- **Train the Matilda model**
- **Multi-task on the training data** — simulation, dimension reduction, feature selection
- **Multi-task on the query data** — classification, dimension reduction, feature selection

Each R call below is annotated with the Python CLI flag it runs underneath. Data
loading is shown explicitly, because the point is to show you *how to run your own data*.

## Requirements

- **R packages**: `matilda`, `SingleCellExperiment` (always); `Seurat` (the
  `.rds` loading option and the latent-space UMAP) and `zellkonverter` (the
  `.h5ad` loading option) — those chunks skip themselves if the package is missing.
- **Python**: nothing to install — `matilda` provisions a private environment via
  `basilisk` on the **first `matilda_train()` call** (a one-off multi-minute build,
  cached afterwards; a GPU is used if present, otherwise CPU).
- **Demo data** (set the two paths below):
  - `demo_h5/`  → `train_rna.h5 train_adt.h5 train_atac.h5 train_cty.csv` and the
    matching `test_*` (8 files);
  - `demo_fmt/` → `train_seurat.rds` and `train_{rna,adt,atac}.h5ad`
    (only needed for loading Options B / C).

```{r deps}
library(matilda)
library(SingleCellExperiment)
demo_h5  <- getOption("matilda.demo",         "~/Documents/matilda-r/teaseq")
demo_fmt <- getOption("matilda.demo_formats", "~/Documents/matilda-r/teaseq_formats")
```

# 1. What Matilda expects

A `SingleCellExperiment` with **RNA raw counts** in the main assay, **ADT** and
**ATAC** as `altExp`s, and a cell-type column in `colData`. ATAC must be a
gene-activity score (gene-level), not raw peaks. (Matilda also accepts a
`MultiAssayExperiment`, a `Seurat` object, or plain matrices.)

# 2. Load your data — pick the format you have

## Option A — native Matilda `.h5` (`matrix/data` / `features` / `barcodes`)

The reading is fully explicit (no hidden wrapper):

```{r load-h5}
read_h5 <- function(path) {
  m     <- rhdf5::h5read(path, "matrix/data")
  feats <- as.character(rhdf5::h5read(path, "matrix/features"))
  cells <- as.character(rhdf5::h5read(path, "matrix/barcodes"))
  if (nrow(m) == length(cells) && ncol(m) == length(feats)) m <- t(m)  # -> genes x cells
  stopifnot(nrow(m) == length(feats), ncol(m) == length(cells))
  dimnames(m) <- list(feats, cells)
  m
}
load_h5_sce <- function(dir, split) {
  g <- function(mod) file.path(dir, sprintf("%s_%s.h5", split, mod))
  sce <- SingleCellExperiment(assays = list(counts = read_h5(g("rna"))))
  altExp(sce, "ADT")  <- SummarizedExperiment(list(counts = read_h5(g("adt"))))
  altExp(sce, "ATAC") <- SummarizedExperiment(list(counts = read_h5(g("atac"))))
  sce$cell_type <- as.character(
    utils::read.csv(file.path(dir, sprintf("%s_cty.csv", split)), header = FALSE)[[2]])[-1]
  sce
}

sce <- load_h5_sce(demo_h5, "train")
sce
lapply(altExps(sce), dim)                 # ADT / ATAC: features x cells
```

This matches the section-1 contract: RNA counts in the main assay, ADT/ATAC as
`altExp`s, labels in `colData$cell_type`.

## Option B — a `Seurat` object

Needs `install.packages("Seurat")`; skip if you only have `.h5`.

```{r load-seurat, eval = requireNamespace("Seurat", quietly = TRUE)}
library(Seurat)
seu <- readRDS(file.path(demo_fmt, "train_seurat.rds"))   # RNA/ADT/ATAC assays + cell_type
ga  <- function(a) as.matrix(SeuratObject::LayerData(seu, assay = a, layer = "counts"))
sce_seurat <- SingleCellExperiment(assays = list(counts = ga("RNA")))
altExp(sce_seurat, "ADT")  <- SummarizedExperiment(list(counts = ga("ADT")))
altExp(sce_seurat, "ATAC") <- SummarizedExperiment(list(counts = ga("ATAC")))
sce_seurat$cell_type <- seu$cell_type

# identical to Option A — every modality + the labels:
c(RNA  = isTRUE(all.equal(as.matrix(counts(sce)),               ga("RNA"))),
  ADT  = isTRUE(all.equal(as.matrix(assay(altExp(sce, "ADT"))),  ga("ADT"))),
  ATAC = isTRUE(all.equal(as.matrix(assay(altExp(sce, "ATAC"))), ga("ATAC"))),
  labels = identical(as.character(sce$cell_type), as.character(sce_seurat$cell_type)))
```

## Option C — `.h5ad` (AnnData / scanpy), one file per modality

Needs `BiocManager::install("zellkonverter")`; skip if you only have `.h5`.

```{r load-h5ad, eval = requireNamespace("zellkonverter", quietly = TRUE)}
library(zellkonverter)
ad <- lapply(c(rna = "rna", adt = "adt", atac = "atac"),
             function(n) readH5AD(file.path(demo_fmt, sprintf("train_%s.h5ad", n))))
# select the counts assay by name (a user's AnnData may also carry logcounts):
am <- function(x) as.matrix(SummarizedExperiment::assay(x, "counts"))
sce_h5ad <- SingleCellExperiment(assays = list(counts = am(ad$rna)))
altExp(sce_h5ad, "ADT")  <- SummarizedExperiment(list(counts = am(ad$adt)))
altExp(sce_h5ad, "ATAC") <- SummarizedExperiment(list(counts = am(ad$atac)))
sce_h5ad$cell_type <- as.character(colData(ad$rna)$cell_type)

c(RNA  = isTRUE(all.equal(as.matrix(counts(sce)),               am(ad$rna))),
  ADT  = isTRUE(all.equal(as.matrix(assay(altExp(sce, "ADT"))),  am(ad$adt))),
  ATAC = isTRUE(all.equal(as.matrix(assay(altExp(sce, "ATAC"))), am(ad$atac))),
  labels = identical(as.character(sce$cell_type), as.character(sce_h5ad$cell_type)))
```

## Option D — plain matrices (no SingleCellExperiment)

```{r load-matrices, eval = FALSE}
# genes x cells matrices + a label vector; returns a model handle directly
fit <- matilda_train(rna = rna_mat, adt = adt_mat, atac = atac_mat, cty = labels)
```

Options **B and C are verified element-identical to Option A** above (every modality
and the labels return `TRUE`), so the downstream results don't depend on which
format you started from; Option D follows the same contract. We continue with `sce`
from Option A.

> **Other formats**: 10x CellRanger output reads via
> `DropletUtils::read10xCounts()` (split feature types into RNA/ADT with
> `splitAltExps()`); ATAC **peaks** must first be turned into a gene-activity
> matrix (e.g. `Signac::GeneActivity()`) before use.

# 3. Train the Matilda model

Mirrors the official `main_matilda_train.py --rna --adt --atac --cty` (defaults:
batch 64, epochs 30, lr 0.02, z_dim 100, seed 1). The model is stored inside the
object, so the rest of the workflow pipes.

```{r train}
sce <- matilda_train(sce, label = "cell_type")   # first call also builds the env (one-off)
matilda_model(sce)
```

# 4. Multi-task on the training data

Mirrors the official *Multi-task on the training data*: simulation, dimension
reduction, and feature selection, all run on the reference (training) object.

## Simulation — `--simulation True --simulation_ct -1 --simulation_num 200`

`celltype = NULL` maps to `--simulation_ct -1` ("simulate all cells"): the whole
training set is passed through the trained VAE, reconstructing every cell with its
original label.

```{r sim-train}
sim <- matilda_simulate(sce, n = 200)
dim(sim)                                   # all training cells, reconstructed
table(sim$label)                           # original cell-type distribution
```

## Dimension reduction — `--dim_reduce True`

```{r reduce-train}
sce <- matilda_reduce(sce)                 # adds reducedDim(sce, "MATILDA")
dim(reducedDim(sce, "MATILDA"))            # cells x z_dim (100)
```

The official `qc/visulize_latent_space.Rmd` visualises this latent space by
building a Seurat object from it, running PCA, and taking a UMAP of the first 10
PCs. We reproduce that pipeline exactly (Seurat's `RunPCA`/`RunUMAP` use
`seed.use = 42` internally, so the layout is reproducible):

```{r umap-train, fig.width = 8, fig.height = 5, eval = requireNamespace("Seurat", quietly = TRUE)}
L <- as(t(reducedDim(sce, "MATILDA")), "CsparseMatrix")    # latent dims x cells
rownames(L) <- paste0("z", seq_len(nrow(L))); colnames(L) <- colnames(sce)
seu_umap <- Seurat::CreateSeuratObject(counts = L)
seu_umap[["RNA"]]$data <- seu_umap[["RNA"]]$counts          # latent as data (no normalization)
seu_umap$celltype <- sce$cell_type
seu_umap <- Seurat::FindVariableFeatures(seu_umap, verbose = FALSE)
seu_umap <- Seurat::ScaleData(seu_umap, verbose = FALSE)
seu_umap <- Seurat::RunPCA(seu_umap, verbose = FALSE)
seu_umap <- Seurat::RunUMAP(seu_umap, dims = 1:10, verbose = FALSE)
Seurat::DimPlot(seu_umap, reduction = "umap", group.by = "celltype")
```

## Feature selection — `--fs True`

```{r fs-train}
markers <- matilda_markers(sce)            # data.frame: celltype / feature / importance
head(markers)
```

# 5. Multi-task on the query data

Mirrors the official *Multi-task on the query data*: classification, dimension
reduction, and feature selection, each run on the held-out query with
`reference = sce` (the `--query True` flag).

```{r load-query}
query <- load_h5_sce(demo_h5, "test")      # 5048 held-out cells
```

## Classification — `--classification True --query True`

```{r classify-query}
query <- matilda_classify(query, reference = sce)   # adds colData$matilda_pred / $matilda_prob
table(query$matilda_pred)                           # predicted cell types
sprintf("query accuracy on this machine = %.4f", mean(query$matilda_pred == query$cell_type))
```

## Dimension reduction — `--dim_reduce True --query True`

```{r reduce-query}
query <- matilda_reduce(query, reference = sce)     # adds reducedDim(query, "MATILDA")
dim(reducedDim(query, "MATILDA"))
```

## Feature selection — `--fs True --query True`

```{r fs-query}
qmarkers <- matilda_markers(query, reference = sce)
head(qmarkers)
```

```{r sessioninfo}
sessionInfo()
```
